### 2018-03-16 paper/2-background.tex

\Excel's build algorithm~\cite{excel_recalc} is significantly different from
\Make. Since it is impossible to construct the full dependency graph upfront in
the presence of dynamic dependencies, \Excel uses the \emph{calculation chain}
produced by the previous build as an approximation to the correct topological
order. During recalculation, \Excel processes cells in this order, but can
\emph{defer recalculation of a cell} by moving it down the chain if a newly
discovered dependency has not yet been rebuilt. We further refer to this
algorithm as \emph{reordering}, and will discuss it in more detail
in~\S\ref{sec-implementation-excel}.

@@

\Shake was developed to solve the issue of dynamic
dependencies~\cite{mitchell2012shake} and it excels at them without sacrificing
the minimality requirement. To demonstrate this, we elaborate the \Make example
from~\S\ref{sec-background-make} by adding the following files whose
dependencies are shown in Fig.~\ref{fig-shake}(a):

\begin{itemize}
    \item \cmd{LICENSE} is an input text file containing the project license.
    \item \cmd{release.txt} is a text file listing all release files. This file
    is produced by concatenating input files \cmd{bins.txt} and \cmd{docs.txt}
    that list all binary and documentation files of the project.
    \item \cmd{release.tar} is the release archive built by executing the
    command \cmd{tar} on the release files.
\end{itemize}

The dependencies of \cmd{release.tar} are not known statically: they are
determined by the content of \cmd{release.txt}, which might not even exist
before the build. Makefiles cannot express such dependencies, requiring ad hoc
workarounds such as \emph{build phases}, which are problematic~\cite{hadrian}.
Here is how one can express the task for producing \cmd{release.tar} using \Shake.

@@

\noindent
We first declare the static dependency on \cmd{release.txt}, then read its
content (a list of files) and depend on it, dynamically. Finally, we specify the
command to produce the resulting archive. Crucially, the archive will only be
rebuilt if one of the dependencies (static or dynamic) has changed. For example,
if we create another documentation file \cmd{README} and add it to
\cmd{docs.txt}, \Shake will appropriately rebuild \cmd{release.txt} and
\cmd{release.tar}, discovering the new dependency, as illustrated in
Fig.~\ref{fig-shake}(b).

@@

\Shake also supports the following \emph{early cutoff optimisation}. When it
executes a task and the result is unchanged from the previous build, it is
unnecessary to execute the dependent tasks, and hence \Shake can stop a build
earlier, as illustrated in Fig.~\ref{fig-cutoff}. Not all build systems support
early cutoff: \Make and \Excel do not, whereas \Shake and \Bazel (introduced
below) do.

>>>>

\Excel's build algorithm~\cite{excel_recalc} is significantly different from
\Make. \Excel uses the \emph{calculation chain}
produced by the previous build as an approximation to the correct topological
order. During recalculation, \Excel processes cells in this order, but can
\emph{defer recalculation of a cell} by moving it down the chain if a newly
discovered dependency has not yet been rebuilt. We refer to this
algorithm as \emph{reordering}, and will discuss it in more detail
in~\S\ref{sec-implementation-excel}.

@@

\Shake was developed to solve the issue of dynamic
dependencies~\cite{mitchell2012shake} without sacrificing
the minimality requirement. Building on the \Make example
from~\S\ref{sec-background-make}, we add the following files whose
dependencies are shown in Fig.~\ref{fig-shake}(a):

\begin{itemize}
    \item \cmd{LICENSE} is an input text file containing the project license.
    \item \cmd{release.txt} is a text file listing all files that should be in the release. This file
    is produced by concatenating input files \cmd{bins.txt} and \cmd{docs.txt}
    that list all binary and documentation files of the project.
    \item \cmd{release.tar} is the release archive built by executing the
    command \cmd{tar} on the release files.
\end{itemize}

The dependencies of \cmd{release.tar} are not known statically: they are
determined by the content of \cmd{release.txt}, which might not even exist
before the build. Makefiles cannot express such dependencies, requiring problematic
workarounds such as \emph{build phases} \cite{hadrian}.
In \Shake we can express the rule for \cmd{release.tar} as:

@@

\noindent
We first declare the static dependency on \cmd{release.txt}, then read its
content (a list of files) and depend on each listed file, dynamically. Finally, we specify the
command to produce the resulting archive. Crucially, the archive will only be
rebuilt if one of the dependencies (static or dynamic) has changed. For example,
if we create another documentation file \cmd{README} and add it to
\cmd{docs.txt}, \Shake will appropriately rebuild \cmd{release.txt} and
\cmd{release.tar}, discovering the new dependency, as illustrated in
Fig.~\ref{fig-shake}(b).

@@

\Shake also supports the \emph{early cutoff optimisation}. When it
executes a task and the result is unchanged from the previous build, it is
unnecessary to execute the dependent tasks, and hence \Shake can stop a build
earlier, as illustrated in Fig.~\ref{fig-cutoff}. Not all build systems support
early cutoff: \Make and \Excel do not, whereas \Shake and \Bazel (introduced
below) do.

### 2018-03-16 paper/3-abstractions.tex
Work on S4

But nor can the task description work for \emph{any} \hs{f}; most task
descriptions (e.g. \hs{sprsh1} in \S\ref{sec-task}) require that \hs{f}
satisfies certain properties, such as \hs{Applicatove} or \hs{Monad}. That is
why \hs{Task} has the ``\hs{c}~\hs{f}~\hs{=>}'' constraint in its type,
expressing that \hs{f} can only be instantiated by types that satisfy the
constraint \hs{c}.

@@

\begin{itemize}
    \item The \hs{ideal} store is \emph{consistent} with the \hs{task}, i.e.
    for all possible non-input keys \hs{k}, the result of recomputing the
    \hs{task} matches the value in the \hs{ideal} store:
    \vspace{-1mm}
    \[
    \hs{compute}~\hs{task}~\hs{ideal}~\hs{k}~\hs{==}~\hs{Just}~\hs{(}\hs{getValue}~\hs{ideal}~\hs{k)}.
    \]
    \item The \hs{result} and \hs{ideal} agree on the output \hs{key}, i.e.:
    \vspace{-1mm}
    \[
    \hs{getValue}~\hs{result}~\hs{key}~\hs{==}~\hs{getValue}~\hs{ideal}~\hs{key}.
    \]
    In other words, we require that the output \hs{key} has the correct value,
    but do not impose any restrictions on intermediate keys, therefore
    permitting shallow cloud builds.
    \item The \hs{store}, \hs{result} and \hs{ideal} agree on all input keys
    \hs{k} that belong to the transitive closure of \hs{key}'s dependencies:
    \[
    \,\,\,\,\,\,\,\hs{getValue}~\hs{store}~\hs{k}~\hs{==}~\hs{getValue}~\hs{result}~\hs{k}~\hs{&&}~\hs{getValue}~\hs{store}~\hs{k}~\hs{==}~\hs{getValue}~\hs{ideal}~\hs{k}
    \]
    This requirement asserts that (i) no inputs were corrupted during the
    build, and (ii)~the \hs{ideal} store has constraints not only on the output,
    but also on the inputs.
\end{itemize}
A build system is \emph{correct} if it produces a correct \hs{result} for any
given \hs{task}, \hs{key} and \hs{store}.
}
\label{def-correct}
\vspace{-2mm}
\rule{\textwidth}{0.4pt}

>>>>

But nor can the task description work for \emph{any} \hs{f}; most task
descriptions (e.g. \hs{sprsh1} in \S\ref{sec-task}) require that \hs{f}
satisfies certain properties, such as \hs{Applicative} or \hs{Monad}. That is
why \hs{Task} has the ``\hs{c}~\hs{f}~\hs{=>}'' constraint in its type,
expressing that \hs{f} can only be instantiated by types that satisfy the
constraint \hs{c}.

@@

\begin{itemize}
    \item The \hs{ideal} store is \emph{consistent} with the \hs{task}, i.e.
    for all possible non-input keys \hs{k}, the result of recomputing the
    \hs{task} matches the value in the \hs{ideal} store:
    \vspace{-1mm}
    \[
    \hs{Just}~\hs{(}\hs{getValue}~\hs{ideal}~\hs{k)}~\hs{==}~\hs{compute}~\hs{task}~\hs{ideal}~\hs{k}.
    \]
    \item The \hs{result} and \hs{ideal} agree on the output \hs{key}, i.e.:
    \vspace{-1mm}
    \[
    \hs{getValue}~\hs{result}~\hs{key}~\hs{==}~\hs{getValue}~\hs{ideal}~\hs{key}.
    \]
    In other words, we require that the output \hs{key} has the correct value,
    but do not impose any restrictions on intermediate keys, therefore
    permitting shallow cloud builds.
    \item The \hs{store}, \hs{result} and \hs{ideal} agree on all input keys
    \hs{k} that belong to the transitive closure of \hs{key}'s dependencies:
    \[
    \,\,\,\,\,\,\,\hs{getValue}~\hs{store}~\hs{k}~\hs{==}~\hs{getValue}~\hs{result}~\hs{k}~\hs{&&}~\hs{getValue}~\hs{store}~\hs{k}~\hs{==}~\hs{getValue}~\hs{ideal}~\hs{k}
    \]
    This requirement asserts that (i) no inputs were corrupted during the
    build, and (ii)~the \hs{ideal} store has constraints not only on the output,
    but also on the inputs.
\end{itemize}
A build system is \emph{correct} if it produces a correct \hs{result} for any
given \hs{task}, \hs{key} and \hs{store}.
}
\label{def-correct}
\vspace{-2mm}
\rule{\textwidth}{0.4pt}
